---
title: "Top Natural Disaster Events in terms of Economic and Population Health Damages
  in the United States from 1950 to 2011"
author: "Gerald Del Norte"
date: "February 9, 2019"
output:
  html_document:
    df_print: paged
---

## Synopsis
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This document involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Data Processing (Downloading, Loading and Cleaning Data)

```{r}

## installs and loads packages(if not yet installed)
if(!require(png)){
  install.packages("png")
  library(png)
}
if(!require(plyr)){
  install.packages("plyr")
  library(plyr)
}

if(!require(dplyr)){
  install.packages("dplyr")
  library(dplyr)
}

if(!require(knitr)){
  install.packages("knitr")
  library(knitr)
}
library(png)
library(plyr)
library(dplyr)
library(knitr) 

```
### Downloads the source file if it is not in the current working directory, proceeds if it is 
```{r}


destfilebz2 = "./repdata_data_StormData.csv.bz2"
fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
if(!file.exists(destfilebz2)){
  download.file(fileURL, destfile="./repdata_data_StormData.csv.bz2", method="auto")
}
```
### Reads and cleans data
```{r}

stormx <- read.csv("./repdata_data_StormData.csv.bz2")
  
Statelist<- list("AK","AL","AR","AZ","CA","CO","CT","DC","DE","FL","GA","GU","HI","IA","ID", "IL","IN","KS","KY","LA","MA","MD","ME","MH","MI","MN","MO","MS","MT","NC","ND","NE","NH","NJ","NM","NV","NY", "OH","OK","OR","PA","PR","PW","RI","SC","SD","TN","TX","UT","VA","VI","VT","WA","WI","WV","WY")

#Removes unrelated rows by only retaining the rows that have state in the STATE column.


storm <- stormx[ stormx$STATE %in% Statelist,]

#Removes columns not needed for this analysis

storm <- select(storm, EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP)



```
### This next section substitutes values on the PROPDMGEXP and CROPDMG EXP to make them numerical, 
### The source of information is from https://rpubs.com/flyingdisc/PROPDMGEXP 
```{r}

#creates a backup csv file and reads the file with strings as characters
write.csv(storm,'storm.csv')
storm<-read.csv("storm.csv",row.names=NULL, stringsAsFactors = FALSE)


##creates replacement dataframe for PROPDMGEXP and CROPDMGEXP
PROPDMGEXP = c("H","h","K","k","M","m","B","b","+","-","?","1","2","3","4","5","6","7","8","0","") 
PROPREPVAL = c("100","100","1000","1000","1000000","1000000","1000000000","1000000000","1","0","0","10","10","10","10","10","10","10","10","10","0")

CROPDMGEXP = c("H","h","K","k","M","m","B","b","+","-","?","1","2","3","4","5","6","7","8","0","") 
CROPREPVAL = c("100","100","1000","1000","1000000","1000000","1000000000","1000000000","1","0","0","10","10","10","10","10","10","10","10","10","0")


#merges the new replacement values with the original dataframe
propexpdf = data.frame(PROPDMGEXP, PROPREPVAL)
storm2 <- merge(x=storm, y=propexpdf, by = "PROPDMGEXP", all.x=TRUE)

cropexpdf = data.frame(CROPDMGEXP, CROPREPVAL)
storm3 <- merge(x=storm2, y=cropexpdf, by = "CROPDMGEXP", all.x=TRUE)

#creates another backup prior to plotting
write.csv(storm3,'storm3.csv')
storm3<-read.csv("storm3.csv",row.names=NULL, stringsAsFactors = TRUE)

```

## Results

### Aggregate and summarizes the top 5 events with most fatalities 
```{r}
  deaths <- aggregate(FATALITIES~EVTYPE, storm3, sum)
  deaths <- deaths[with(deaths, order(-FATALITIES)), ]
  deaths <- deaths[1:5,]
  head(deaths)
  png("plot1.png", width=800, height=600)
  plot1 <- barplot(deaths$FATALITIES, names = deaths$EVTYPE, xlab = "Events", ylab = "Fatalities", main = "Deaths by Event Type")
  dev.off() 
  img1 <- readPNG("./plot1.png")
  grid::grid.raster(img1)
  
```

### Aggregate and summarizes the top 5 events with most injuries 

```{r}  
  injur <- aggregate(INJURIES~EVTYPE, storm3, sum)
  injur <- injur[with(injur, order(-INJURIES)), ]
  injur <- injur[1:5,]
  head(injur)
  png("plot2.png", width=800, height=600)
  plot2 <- barplot(injur$INJURIES, names = injur$EVTYPE, xlab = "Events", ylab = "Injuries", main = "Injuries by Event Type")
  dev.off() 
  img2 <- readPNG("./plot2.png")
  grid::grid.raster(img2)
  
# Set factors as numeric  
storm3[,"CROPDMG"] <- as.numeric(as.character(storm3[,"CROPDMG"]))
storm3[,"CROPREPVAL"] <- as.numeric(as.character(storm3[,"CROPREPVAL"]))
storm3[,"PROPDMG"] <- as.numeric(as.character(storm3[,"PROPDMG"]))
storm3[,"PROPREPVAL"] <- as.numeric(as.character(storm3[,"PROPREPVAL"]))

```

## Mutates the dataframe by using the exponents for crop and property damage, then adds the two to determine economic cost.  
```{r}
storm4 <- mutate(storm3, CROPTOTAL = CROPREPVAL * CROPDMG)
storm5 <- mutate(storm4, PROPTOTAL = PROPREPVAL * PROPDMG)
storm6 <- mutate(storm5, COSTTOTAL = CROPTOTAL + PROPTOTAL)

cost <- aggregate(COSTTOTAL~EVTYPE, storm6, sum)
cost <- cost[with(cost, order(-COSTTOTAL)), ]
cost <- cost[1:5,]
head(cost)
png("plot3.png", width=800, height=600)
plot3 <- barplot(cost$COSTTOTAL, names = cost$EVTYPE, xlab = "Type of Event", ylab = "Total Economic Damage", main = "Total Damage Costs by Event Type")
dev.off() 
img3 <- readPNG("./plot3.png")
grid::grid.raster(img3)

```